Complexity-stability relationships in disordered dynamical systems

### 2024-03-29 papers/submission/beyond-may.tex

%Random interactions: equilibrium distribution
\paragraph*{}
\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line was plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation~\eqref{dynamics} a dynamical mean field theory (DMFT) 
which describes the ensemble of $N\gg 1$ variables
by means of a single, self-consistent representative stochastic trajectory
\begin{equation}
    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \tilde{\mu} \langle x^{\gamma}\rangle + \tilde{\sigma} \eta\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\tilde{\mu}=\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
At equilibrium, we can write
\begin{equation}
    x_*^{\alpha-\beta} - A_s x_*^{\gamma}= \big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\end{equation} 
where $\xi$ is now a standard normal random variable.
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\tilde{\sigma}^2 \langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \tilde{\mu} \langle x_*^{\gamma}\rangle)^2}{2\tilde{\sigma}^2\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
where the expectation values must be computed self-consistently.

>>>>

%Random interactions: equilibrium distribution
\paragraph*{}
\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed using the cavity solution and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line is plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution which describes the ensemble of the stationary value of $N\gg 1$ variables by means of a single representative radnom variable $x_*$
\begin{equation}
    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\xi$ is a standard normal random variable. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\tilde{\mu}=\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\tilde{\sigma}^2 \langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \tilde{\mu} \langle x_*^{\gamma}\rangle)^2}{2\tilde{\sigma}^2\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
where the expectation values must be computed self-consistently.
